Natural Selection Across Three Generations of Americans

### 2024-03-03 response-to-reviewers-revision-1.qmd

We now detail the theory more in the abstract, and we've described the
substitution effects and income effects straight away in the intro.
We've tried to keep the language as simple and intuitive as possible.

*I was wondering why the authors control for PCAs in the PGI analyses.
As stated by the authors, this study is not a causal study but one of
correlations. The consequence of a correlation (natural selection)
should be independent of its causes.*

Why residualize or control for principal components at all? I think for
two reasons. First, we are ultimately interested in selection on the
"true PGI". Stripping out PCs is an admittedly imperfect way to control
for confounding population stratification. Second, PGIs are typically
residualized in this way, so by examining the effect of these we're
following the "industry standard" and learning about scores in the form
they are actually used.

To simplify interpretation, we've decided to residualise the PGIs on the
principal components and then standardize them, instead of control for
them.

### 2024-03-04 hrs-selection.qmd
Tweaks before resubmission: - edit response letter - foreground parents' education correlation

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  natural selection: scores which correlate negatively (positively) with
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  half borne out: coefficients are larger only among low-income
  parents and unmarried parents. Using respondents' number of 
  siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

rand <- svydesign(
    id = ~ raehsamp,
		strata = ~ raestrat,
		weights = ~ weight, 
		nest = TRUE,
		data = rand |> drop_na(weight) 
		)

@@

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The motivation for our study is to establish the direction and magnitude of natural selection upon a range of traits in Americans and to test whether the economic theory can explain the selection. Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans. To preview our results, we confirm point 1 above, but only see partial and ambiguous support for point 2. (TODO: or can we be stronger?)

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants. We focus on these because the sample size is large enough. The appendix reports the most basic analyses for the `r n_black` genotyped black participants.

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants. PGS were residualised on the first ten within-ethnicity principal components of the DNA array data, to reduce bias from population stratification. PGS were then rescaled to zero mean and unit variance.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated separately by ethnicity. This is not ideal, because it treats the ethnicities as separate breeding populations. But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative. We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights.

The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection. However the approach is also based on quantitative genetic theory and is able to yield an estimate of genetic change from one generation to the next.

The Robertson-Price Identity states that the change in a genetic trait between one generation and the next is equal to its covariance with relative fitness [@Price1970identity; @Robertson1966identity], assuming no mutations. In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness. As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score per generation. Because the polygenic scores are standardized to have a variance of 1, the regression coefficient of RLRS on the PGS is equal to the covariance. Our approach to measuring natural selection is not novel, having been employed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].

@@

```{r}
#| label: stat-regressions

# cache seems to cause trouble even with cache-lazy: FALSE 
options("survey.lonely.psu" = "average")

@@

We estimate coefficients of PGS on RLRS. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, self-rated health and having ever smoked. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   Dashed lines are fitted from linear regressions.

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not perfectly comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are significant and negative (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (`r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

The HRS has data on the years of education for the parents of respondents. We recalculate the coefficients on education using values for the parents, in case the genetics of education and human capital changed between generations. We then correlate these new coefficients with the number of siblings, finding a positive correlation again (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`). The estimate is slightly smaller as would be expected since respondents likely recall their parent's education with some error.

@@

Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education. Correlations are significant. Standard errors are large (correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`). There is a positive correlation between PGS coefficients on RLRS2 and RLRS (`r gen_cor_gkids`, bootstrapped 95% C.I. `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.

@@

We should attain more statistical power if we utilise information from other polygenic scores at the same time. However, some caution is necessary. Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate. We test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

@@

where $DIS_i$ is a dummy for $i$'s membership in the disadvantaged group and where $PGS_i$ has been sign-flipped to correlate negatively with education. Under the null, the estimated $\delta$s will be distributed around zero; in expectation half will be positive and half negative. Under the alternative, more than half the $\delta$s will be positive. We bootstrap respondents and count the number of positive $\delta$s in each resample. If 95% confidence intervals are above `r length(pgs_white)`/2 = `r length(pgs_white)/2`, we reject the null in favour of the alternative. \[TODO: what if some are negative and some are positive? I think this is fine because it makes our test conservative? Maybe??\]

@@

To complement the above cross-PGS analysis, we focus on two scores which we a priori expect to show intergroup differences: educational attainment and ADHD. Both of these showed large differences in @hugh2022human, and have significant main effects in this data. Figure @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups. All differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA. However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction. \[XXX or we just say "nothing survives B. C.\]

@@

tidy_coefs |> 
  # Select only the interaction terms:
  filter(grepl(":", term)) |> 
  mutate(.by = c(pgs, interaction),
    PGS = pgs,
    # gets the pretty name of the group
    Split = names(interactions)[match(interaction, interactions)],
    Group = c("Disadvantaged", "Advantaged"),
    Difference = ifelse(test.p.value < 0.05, "Difference p < 0.05",
                        "Difference n.s.")
  ) |> 
  ggplot(aes(estimate, Split, colour = Group)) + 
    geom_linerange(aes(xmin = conf.low, xmax = conf.high), alpha = 0.5,
                   position = pos_dod) +
    geom_point(aes(shape = Group, colour = Group, fill = Difference),
               position = pos_dod, 
               shape = "circle filled", size = 2) +
    facet_wrap(vars(PGS), ncol = 1) +
    scale_fill_manual(values = c("Difference n.s." = "white", 
                                 "Difference p < 0.05" = "orange"),
                                 guide = guide_legend(title = NULL)) +
    scale_color_manual(values = c("Advantaged" = "steelblue3", 
                                  "Disadvantaged" = "red4")) +
    theme_minimal() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank()
    ) + 
    labs(x = "Coefficient on RLRS", y = "")

@@

@@

@fig-true-effects reports the error corrected estimates of selection. As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations. For twin-heritability, different relationships with fertility among variants not measured on the chip may also bias the estimates. Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.

To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits. Given the focus on cognitive traits, it is interesting to see that the point estimates for many other traits exceed point estimates for educational attainment and cognitive performance. Depressive symptoms, asthma, age of first birth and self-rated health all appear to be changing faster. This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also of medical relevance.

To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability. For cognitive performance we have estimated a genetic change of $`r IQgen_change`$ standard deviations per generation, assuming a twin heritability of $0.51$. This implies a phenotypic change of $`r IQgen_change`\times0.51^\frac{1}{2}=`r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points. This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype. Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.

Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century. Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased in the late twentieth century, contrary to how we believe the genetics have changed. However, the prevalence of self-reported anxiety and depressive symptoms have increased [@Witters2023; @Twenge2000], paralleling the polygenic scores for depression and neuroticism. The prevalence of asthma has also increased [@Moorman2007] in line with the polygenic score. The apparent paradox of phenotypes and nautral selection moving in opposite directions suggests that environmental change across history is a much greater force than that of natural selection.

@@

@@

The economic theory of fertility is driven by the trade-off between children and income. On the one hand, a higher hourly wage increases the opportunity cost of raising children (a "substitution effect"). On the other hand, higher expected income makes children more affordable (an "income effect"). Prediction 1 of the theory holds when substitution effects dominate income effects overall. Prediction 2 is driven by the specific form of individuals' preferences for income: when utility for income is sharply curved, i.e. marginal utility is high at low incomes but low at high incomes, the substitution effect is stronger for those who expect to earn less.

The results here support prediction 1 but are more ambiguous for prediction 2. PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents. But there is little evidence for larger coefficients among people with lower education, or younger parents. This may be due to the low sample size. But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here. The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason. But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^3] In this sense, results here are less supportive of the economic theory specifically.

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated. But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.

Across three generations of Americans, we found estimates of selection to correlate over time and that selection is always more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The result is congruent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929. The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is slightly smaller than in previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to their social implications. To know more, we must await more accurate polygenic scores.

@@

```{r}
#| label: fig-siblings
#| fig-cap: |
#|   Scatterplot of PGS coefficients on number of live siblings and years of
#|   education. Dashed lines show linear regressions.

tidy_pgs_sibs$white |> 
  inner_join(tidy_pgs_ed$white, 
             by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
  ggplot(aes(estimate.edyrs, estimate.sibs)) +
    geom_hline(yintercept = 0, color = "grey80") +
    geom_vline(xintercept = 0, color = "grey80") +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                linetype = "dashed", colour = "red", linewidth = 0.5) +
    coord_cartesian(xlim = c(-.3, .75)) +
    theme_light() + 
    theme(panel.grid = element_blank()) +
    labs(
        x = "Coefficient on years of education",
        y = "Coefficient on number of siblings"
    )

@@

```{r}
#| label: fig-gkids
#| fig-cap: |
#|   Scatterplot of PGS coefficients on RLRS2 and years of
#|   education. RLRS2 is the reproductive success of respondent's offspring
#|   relative to the success of other offspring. Dashed lines show linear regressions.

@@

```{r}
#| label: fig-check-correction
#| eval: false
#| fig-cap: |
#|   Estimated coefficients of series of PGS on fertility 
#|   using errors-in-variables correction. "Own" is the uncorrected estimate
#|   using the PGS itself. Other estimates are from earlier PGS targeting
#|   the same phenotype, corrected via the method in @becker2021resource.
#| fig-height: 6

corrected_effects |> 
  filter(length(unique(source.pgs)) > 1, .by = phenotype) |> 
  filter(! target.pgs %in% c("Chip", "Twin")) |> 
  filter(
    ! target.pgs %in% old_pgs
  ) |> 
  mutate(
    Phenotype = case_match(phenotype,
                           "adhd_score" ~ "ADHD",
                           "r10cesd" ~ "Depression",
                           "r10cogtot" ~ "Cognition",
                           "raedyrs" ~ "Educ. attainment"
                           ),
    Target = target.pgs,
    Source = ifelse(source.pgs == target.pgs, "Own", source.pgs),
  ) |> 
  ggplot(aes(y = Target, x = mean, xmin = conf.low, xmax = conf.high,
             color = Source)) +
    geom_vline(xintercept = 0, color = "grey60") +
    geom_pointrange(position = position_dodge(0.7)) +
    geom_text(aes(x = mean, label = Source), hjust = "center", 
              vjust = 1.5,
              position = position_dodge(0.7), size = 2.5) + 
    scale_color_manual(values = c("Own" = "red"), guide = "none") +
    facet_grid(rows = vars(Phenotype), scales = "free", space = "free") +
    labs(x = "Corrected coefficient on RLRS", y = "Target PGS") +
    coord_cartesian(xlim = c(-0.05, 0.05)) + 
    theme_minimal()

>>>>

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Across
  three generations, scores which correlate negatively (positively) with
  education are selected for (against). However, results only
  partially support the economic theory of fertility as an explanation for 
  natural selection. The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  half borne out: coefficients are larger only among low-income
  parents and unmarried parents. We also estimate effect sizes corrected 
  for noise in the polygenic scores. Selection for some health traits is 
  similar in magnitude to that for cognitive traits. 
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

rand <- svydesign(
    id = ~ raehsamp,
		strata = ~ raestrat,
		weights = ~ weight, 
		nest = TRUE,
		data = rand |> drop_na(weight))

@@

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The motivation is to establish the direction and magnitude of natural selection on a range of traits in Americans, and to test whether the economic theory can explain the selection. Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans. This is interesting first because selection effects may accumulate over time, and second because an possible alternative explanation for existing findings is the development of the welfare state, which happened in the US during the "Great Society" programs of the 1960s (i.e. after the respondents' parents' generation). To preview our results, we confirm point 1 above across all three generations. But we only see partial and ambiguous support for point 2.

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. We include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants. We focus on these because the sample size is large enough. The appendix reports some basic analyses for the `r n_black` genotyped black participants.

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. We also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants. PGS were residualised on the first ten within-ethnicity principal components of the DNA array data, to reduce bias from population stratification. PGS were then rescaled to zero mean and unit variance.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated separately by ethnicity. This is not ideal, because it treats the ethnicities as separate breeding populations. But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative. We therefore focus on the white US population, with the caveat that results from this "one data point" may not replicate in other ethnicities or countries. The mean number of children of people born in each year was calculated using sampling weights.

The intuition behind our analysis is simple: if a polygenic score predicts more reproductive success, then people higher in the PGS will reproduce more than others, causing scores to increase in the population. However the approach is also based on quantitative genetic theory and is able to yield an estimate of genetic change from one generation to the next.

The Robertson-Price Identity states that the change in a genetic trait between one generation and the next is equal to its covariance with relative fitness [@Price1970identity; @Robertson1966identity], assuming no mutations. In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness. As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score per generation. Because the polygenic scores are standardized to have a variance of 1, the regression coefficient of RLRS on the PGS is equal to the covariance. This approach is standard in the literature [@Beauchamp2016selection; @hugh2022human; @kong2017selection].

@@

```{r}
#| label: stat-regressions
# cache seems to cause trouble even with cache-lazy: FALSE

options("survey.lonely.psu" = "average")

@@

We estimate coefficients of PGS on RLRS. These do not identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, self-rated health and having ever smoked. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   The dashed line shows a linear regression.

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not perfectly comparable across the generations. The HRS has data on the years of education for the parents of respondents. We recalculate the coefficients on education using values for the parents, in case the genetics of education and human capital changed between generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on parents' years of education. The correlation is negative and significant (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`). The same holds if we use coefficients on own years of education (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (`r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

@@

Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education. Correlations are significant. Standard errors are large (correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`). Again, there is a positive correlation between PGS coefficients on RLRS2 and RLRS (`r gen_cor_gkids`, bootstrapped 95% C.I. `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating stability in selection over time.

@@

We test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

@@

where $DIS_i$ is a dummy for $i$'s membership in the disadvantaged group and where $PGS_i$ has been sign-flipped to correlate negatively with education. Under the null, the estimated $\delta$s will be distributed around zero; in expectation half will be positive and half negative. Under the alternative, more than half the $\delta$s will be positive. We bootstrap respondents and count the number of positive $\delta$s in each resample. If 95% confidence intervals are above `r length(pgs_white)`/2 = `r length(pgs_white)/2`, we reject the null in favour of the alternative.

@@

To complement the above cross-PGS analysis, we focus on two scores which we a priori expect to show intergroup differences: educational attainment and ADHD. Both of these showed large differences in @hugh2022human, and have significant main effects in this data. @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups. 7 out of 8 differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA. However, no differences remain significant after Bonferroni correction.

@@

tidy_coefs |> 
  # Select only the interaction terms:
  filter(grepl(":", term)) |> 
  mutate(.by = c(pgs, interaction),
    PGS = pgs,
    # gets the pretty name of the group
    Split = names(interactions)[match(interaction, interactions)],
    Group = c("Disadvantaged group", "Advantaged group"),
    Difference = ifelse(test.p.value < 0.05, "Difference p < 0.05",
                        "Difference n.s.")
  ) |> 
  ggplot(aes(estimate, Split, colour = Group)) + 
    geom_linerange(aes(xmin = conf.low, xmax = conf.high), alpha = 0.5,
                   position = pos_dod) +
    geom_point(aes(shape = Group, colour = Group, fill = Difference),
               position = pos_dod, 
               shape = "circle filled", size = 2) +
    facet_wrap(vars(PGS), ncol = 1) +
    scale_fill_manual(
      values = c("Difference n.s."     = "white", 
                 "Difference p < 0.05" = "orange"),
      guide = guide_legend(title = NULL)) +
    scale_color_manual(
      values = c("Advantaged group"    = "steelblue3", 
                 "Disadvantaged group" = "red4"),
      guide = guide_legend(title = NULL)) +
    theme_minimal() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank()
    ) + 
    labs(x = "Coefficient on RLRS", y = "")

@@

## Corrected effect sizes

@@

@fig-true-effects reports the error corrected estimates of selection. As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "substantive": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations. For twin-heritability, different relationships with fertility among variants not measured on the chip may also bias the estimates. Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.

To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits. The point estimates for many traits, including depressive symptoms, asthma, age of first birth and self-rated health, exceed point estimates for educational attainment and cognitive performance. This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also relevant to population health.

To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability. For cognitive performance we have estimated a genetic change of $`r IQgen_change`$ standard deviations per generation, assuming a twin heritability of $0.51$. This implies a phenotypic change of $`r IQgen_change` \times 0.51^\frac{1}{2} = `r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points. This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype. Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.

Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century. Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased in the late twentieth century, contrary to how we believe the genetics have changed. However, the prevalence of self-reported anxiety and depressive symptoms has increased [@Witters2023; @Twenge2000], paralleling the polygenic scores for depression and neuroticism. The prevalence of asthma has also increased [@Moorman2007] in line with the polygenic score. The apparent paradox of phenotypes and natural selection moving in opposite directions suggests that environmental change is a much stronger force than that of natural selection across recent history.

@@

# Now we just do this using the python tool

@@

Across three generations of Americans, estimates of selection correlate over time, and selection is always more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. This result fits that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a sample of Swedish twins. The underlying causes of selection seem to have stayed the same over time, so that the effects of selection have accumulated. In particular, selection against traits associated with education clearly predates the development of the US welfare state.

The economic theory of fertility is driven by the trade-off between children and income. On the one hand, a higher hourly wage increases the opportunity cost of raising children (a "substitution effect"). On the other hand, higher expected income makes children more affordable (an "income effect"). Prediction 1 of the theory holds when substitution effects dominate income effects overall. Prediction 2 is driven by the specific form of individuals' preferences for income: when marginal utility is higher at lower income levels, the substitution effect is stronger for those who expect to earn less.

The results here support prediction 1 but are more ambiguous for prediction 2. PGS coefficients on RLRS appear larger for low-income groups and unmarried respondents. But there is little evidence for larger coefficients among people with lower education, or younger parents. This may be due to the low sample size. But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here. The theory can accommodate this non-result, if preferences for income are less curved in the US for any reason. But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^3] In this sense, results here are less supportive of the economic theory specifically.

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that in the US, SES maps on to race, so that ethnic differences capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents [blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority; @johnson1979minority; @yang2003big; @maralani2013demography]. And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated. But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Another possibility is that the US cohort was exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is slightly smaller than in previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to its social implications. To know more, we must await more accurate polygenic scores.

@@

```{r}
#| label: fig-siblings
#| fig-cap: |
#|   Scatterplot of PGS coefficients on number of live siblings and years of
#|   education for parents. The dashed line shows a linear regression.

tidy_pgs_sibs$white |> 
  inner_join(tidy_pgs_ed_parent$white, 
             by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
  ggplot(aes(estimate.edyrs, estimate.sibs)) +
    geom_hline(yintercept = 0, color = "grey80") +
    geom_vline(xintercept = 0, color = "grey80") +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                linetype = "dashed", colour = "red", linewidth = 0.5) +
    coord_cartesian(xlim = c(-.3, .75)) +
    theme_light() + 
    theme(panel.grid = element_blank()) +
    labs(
        x = "Coefficient on years of education",
        y = "Coefficient on number of siblings"
    )

@@

```{r}
#| label: fig-gkids
#| fig-cap: |
#|   Scatterplot of PGS coefficients on RLRS2 and years of
#|   education. RLRS2 is the reproductive success of respondent's offspring
#|   relative to the success of other offspring. The dashed line shows a 
#|   linear regression.

@@
